---
title: "XCNV"
output:
  github_document:
    toc: true
    html_preview: false
---

<!-- README.md is rendered from this file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# XCNV

XCNV annotates copy-number variants and evaluates the published X-CNV
classifier without bedtools or a runtime XGBoost dependency. It also provides a
separate ACMG/ClinGen 2019 lane in which every selected criterion remains a row
that can be inspected, changed, and rescored.

The implementation is table-oriented. DuckDB evaluates the genomic overlap and
reciprocal-overlap joins; ordinary R owns input validation, feature reduction,
the portable tree evaluator, and evidence scoring.

## Installation

```r
install.packages(
  "XCNV",
  repos = c(
    "https://rgenomicsetl.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

## A real ClinVar deletion

The package includes a compact, case-limited annotation set for ClinVar allele
72461: a pathogenic deletion containing `SOX2`. This is a real scientific
example, not the synthetic fixture used by the test suite.

```{r real-case}
library(XCNV)

example <- xcnv_real_example()
example$clinvar[, c(
  "clinvar_allele_id", "clinical_significance",
  "grch38_chrom", "grch38_start1", "grch38_end1",
  "grch37_chrom", "grch37_start0", "grch37_end0"
)]
```

ClinVar publishes this record in both assemblies. X-CNV uses its GRCh37
half-open interval. When a source provides only one assembly, lift it during
resource staging—for example with Rduckhts' `rduckhts_liftover()`—and retain
the source and destination coordinates beside the case.

The following call evaluates all five X-CNV annotation families and the bundled
published model:

```{r real-prediction}
prediction <- predict_cnv(example$cnv, example$resources)
prediction[, c(
  "Chr", "Start", "End", "Type",
  "CDTS_1st", "CDTS_5th",
  "gain.freq", "loss.freq",
  "pLI", "Episcore", "GHIS", "MVP_score"
)]
```

`MVP_score` is the X-CNV model output; XCNV does not relabel it as an ACMG
criterion or clinical class.

## Inspectable ACMG/ClinGen evidence

The same real deletion completely contains the GENCODE v19 `SOX2` gene and its
ClinGen haploinsufficiency record. The automatic provider derives only the
objective content criteria it can establish from those relations:

```{r real-evidence}
cnv <- transform(
  example$cnv,
  cnv_id = example$clinvar$cnv_id,
  cnv_type = type
)
evidence <- derive_cnv_2019_evidence(
  cnv, example$genes, example$dosage
)
evidence[, c(
  "criterion", "score", "source_id",
  "source_release", "evidence_id"
)]
```

The rows can be scored directly:

```{r real-classification}
score_cnv_2019(evidence)
```

Here criterion 2A supplies one point because the deletion completely contains
an established haploinsufficient gene. The automatic provider does not invent
phenotype, inheritance, segregation, population, case-report, or partial-gene
evidence. Those require explicit additional evidence rows.

## Published model without XGBoost

The published classifier is one tree with 41 nodes. XCNV stores the tree as a
small, validated tabular object and evaluates it directly:

```{r model}
model <- xcnv_bundled_model()
data.frame(
  trees = length(unique(model$nodes$tree_id)),
  nodes = nrow(model$nodes),
  features = length(model$feature_names)
)

model_data <- new.env(parent = emptyenv())
data("xcnv_portable_model", package = "XCNV", envir = model_data)
identical(model_data$xcnv_portable_model, model)
```

`tools/stage_xcnv_model.R` rebuilds this portable form from the original
serialized booster and compares both evaluators on 1,024 deterministic feature
rows. XGBoost is needed for that maintainer-only conversion, not for package
installation or prediction.

## Full annotation resources

The bundled real example is deliberately valid only for ClinVar allele 72461.
A general run uses a full GRCh37 resource directory:

```r
resources <- xcnv_resources("/path/to/xcnv-resources", require = "annotations")
cnvs <- read_cnv("input.bed")
predictions <- predict_cnv(cnvs, resources)
write_xcnv(predictions, "predictions.output.csv")
```

`xcnv_resource_manifest()` lists the required files. `run_xcnv()` provides the
same workflow as one call and writes only when an output path is supplied.

The compatibility contract, coordinate conventions, and current gaps are in
[`docs/upstream-compatibility.md`](https://github.com/RGenomicsETL/XCNV/blob/main/docs/upstream-compatibility.md).

## Citation

Please cite Zhang et al. (2021), *Genome Medicine* 13:132,
[doi:10.1186/s13073-021-00945-4](https://doi.org/10.1186/s13073-021-00945-4).
